---
title: "biobouncer: a gate for biological inputs"
subtitle: "Validate gene symbols, ontology terms and accessions before they reach a tool"
---

```{r}
#| label: setup
#| message: false
source(here::here("R", "demo-env.R"))
library(biobouncer)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
packageVersion("biobouncer")
```

Every bioinformatics app I have written had a function somewhere called
`is_gene()` or `looks_like_mondo()`. Each one was a regex written in a hurry.
Each one was slightly different from the one in the previous app. None of
them knew that `MLL` was renamed `KMT2A` in 2013.

biobouncer is one small API that answers "is this an identifier, and if not,
what did the user probably mean?" for 50 sources. It gives the same answer in
R, Python and JavaScript because all three are tested against one shared
corpus. It does not fetch annotation. It runs before anything that does.

## 1. The regex you already have

Here is the kind of check that lives in most apps. It does two things wrong
in one line.

```{r}
is_gene <- function(x) grepl("^[A-Z0-9-]+$", x)
is_gene(c("MLL", "KMT2A", "mondo:5148"))
```

`MLL` passes, and it is a retired symbol. `mondo:5148` fails, and it is one
uppercase and three zeros away from a real disease id. The regex knows the
shape and nothing else.

## 2. Discover what can be checked

```{r}
srcs <- sources()
length(srcs)
head(srcs, 12)
```

`source_info()` is the registry as a table: a valid example for each source,
the modes it supports, and whether it is species or version aware. You never
have to invent an example id.

```{r}
info <- source_info()
info[info$key %in% c("hgnc", "mondo", "go", "ensembl", "uniprot", "hgvs"), ]
```

Four modes, from cheapest to most expensive:

| mode | question | network |
|---|---|---|
| `pattern` | is the string well formed? | no |
| `cache` | does the id exist in a pinned snapshot? | no |
| `remote` | does the id exist right now, in the live source? | yes |
| `existence` | snapshot first, remote only if there is no snapshot | maybe |

## 3. Snapshots are dated

`cache` mode checks existence against a snapshot. The package bundles a full
HGNC snapshot and small samples for a few ontologies; everything else is one
`biobouncer_pull()` away.

```{r}
biobouncer_snapshots()
```

The `version` column is the point. A snapshot has a date. When you write down
which one you checked against, someone can re-run your check in a year and
get the same verdict.

## 4. Mode `pattern`: shape only, offline

```{r}
check_id(c("MONDO:0005148", "mondo:5148", "GO:0006915"), source_db = "mondo")
```

Three things to notice in that table. The lowercase id is invalid but carries
a `suggestion`, which is what a user meant. The GO term is well formed and is
still rejected, because it is not a MONDO id. And the result has one row per
input, in input order, which is what lets it sit next to your data frame.

## 5. Mode `cache`: existence, and repair

A real, messy association table. Retired gene symbols, an obsolete GO term, a
mis-cased disease id, an id that does not exist, and a blank.

```{r}
df <- read.csv("data/associations.csv", stringsAsFactors = FALSE, na.strings = "")
df
```

`report_id()` runs `check_id()` on a column and prints a one line summary.

```{r}
cols <- c(gene = "hgnc", disease = "mondo", process = "go")
for (col in names(cols)) {
  print(report_id(df[[col]], cols[[col]], how = "cache"))
}
```

Look inside the gene report. `MLL` and `CARS` are retired, and the snapshot
knows their successors.

```{r}
check_id(df$gene, source_db = "hgnc", how = "cache")
```

`repair_id()` substitutes every fixable value and leaves the rest alone, so
it drops into `dplyr::mutate()`.

```{r}
clean <- df
clean$gene <- repair_id(df$gene, "hgnc", how = "cache")
clean$disease <- repair_id(df$disease, "mondo", how = "cache")
clean$process <- repair_id(df$process, "go", how = "cache")
clean
```

`notagene` and `MONDO:9999999` are still there, unchanged, because there is
nothing honest to replace them with. Repair never invents.

## 6. `pattern` across many sources at once

A second table mixes accessions from six databases. Each row is checked
against its own source.

```{r}
ids <- read.csv("data/identifiers.csv", stringsAsFactors = FALSE)
res <- do.call(rbind, lapply(split(ids, ids$source_db), function(g) {
  check_id(g$value, source_db = g$source_db[1], how = "pattern")
}))
res[, c("input", "source_db", "valid", "normalized", "suggestion")]
```

One call shape, 50 sources. The HGVS check is a real grammar, not a regex:
`c.bad` fails for a reason, and so would `c.76insG`, which is missing the
second position an insertion needs.

## 7. Mode `remote`: live, and never a crash

`remote` asks the source API. With `on_error = "indeterminate"`, an id whose
check could not complete comes back as an `NA` row with the reason in the
`error` column, and the rest of the batch still gets checked. No `tryCatch()`
at the call site. Offline, this chunk still renders.

```{r}
check_id(
  "ENSG00000139618",
  source_db = "ensembl",
  how = "remote",
  species = "homo_sapiens",
  on_error = "indeterminate"
)
```

`remote` is also species aware. The same Ensembl id under `mus_musculus`
would be rejected in `pattern` mode before any network call, because mouse
ids start with `ENSMUSG`.

## 8. Pin it

The reproducible workflow is: pull a snapshot once, commit the version string
to your methods, run `cache` mode forever after.

```{r}
#| eval: false
biobouncer_pull("mondo")
biobouncer_snapshots()
```

Then `check_id(x, "mondo", how = "cache", version = "2026-09-01")` names
the exact release you validated against.

## 9. Drops into your validation stack

The same verdict through every door. checkmate style:

```{r}
check_valid_id(c("MONDO:0005148", "mondo:5148"), "mondo")
test_valid_id(c("MONDO:0005148", "MONDO:0018076"), "mondo")
```

`assert_valid_id()` throws, which is what you want at the top of a function.

```{r}
#| error: true
assert_valid_id(c("MONDO:0005148", "mondo:5148"), "mondo")
```

An element-wise predicate for assertr, validate, pointblank or a plain
`Filter()`:

```{r}
is_hgnc <- id_predicate("hgnc", how = "cache")
is_hgnc(df$gene)
df[is_hgnc(df$gene), ]
```

And a shinyvalidate rule, so a Shiny input cannot fire a lookup on a
malformed id. The Shiny app in this repo uses this one.

```{r}
rule <- sv_biobouncer("hgnc", how = "cache")
rule("TP53")
rule("notagene")
```

## 10. Same answer in Python

The Python package is tested against the same corpus, so the verdicts match.
This section runs through reticulate; if Python biobouncer is not available
the chunks are skipped and the document still renders.

```{r}
#| label: py-setup
#| message: false
has_py <- tryCatch(
  {
    reticulate::py_require("biobouncer==0.2.0", python_version = "3.12")
    reticulate::py_module_available("biobouncer")
  },
  error = function(e) FALSE
)
has_py
```

```{python}
#| eval: !expr has_py
import biobouncer as bb
bb.__version__
for r in bb.check_id(["MONDO:0005148", "mondo:5148", "GO:0006915"], "mondo"):
    print(r.input, r.valid, r.normalized, r.suggestion)
```

The same gene column, the same repair.

```{python}
#| eval: !expr has_py
import csv
with open("data/associations.csv", newline="") as f:
    genes = [row["gene"] or None for row in csv.DictReader(f)]

rep = bb.report(genes, "hgnc", how="cache")
print(rep.summary)
print(rep.repair())
```

Compare with the R column from section 5: same rows, same suggestions, same
two values left alone.

```{r}
clean$gene
```

## What to take away

- Validate before you fetch. A network call on a malformed id is a wasted
  call, and a network call on a retired symbol is a wrong answer.
- `pattern` and `cache` are offline and deterministic. Put them in tests.
- Snapshots have versions. Write the version down.
- Repair, do not drop. A user who typed `MLL` meant `KMT2A`.
- One API, three languages, one corpus.

Docs: <https://www.samuelbharti.com/biobouncer/>. Install with
`install.packages("biobouncer")` or `pip install biobouncer`.
