---
title: "bioclients: 29 services, one shape"
subtitle: "Look up genes, variants and proteins without writing a client per service"
---

```{r}
#| label: setup
#| message: false
source(here::here("R", "demo-env.R"))
library(bioclients)
library(biohttp)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
packageVersion("bioclients")
```

Four apps in the same family had four gnomAD clients. They disagreed with
each other about chunk sizes, about what a missing gene looks like, and about
what happens when the service is down. bioclients is those clients written
once, on top of biohttp, so every one of them returns the same envelope and
the same shape of table.

Every call here is served from the disk cache committed with this repo, so
the document renders with the network off. See the biohttp tutorial for why
that works.

## 1. Installing

bioclients is not on CRAN yet. The package lives in the `pkg-r` folder of
its repo, and that suffix is mandatory.

```r
pak::pak("samuelbharti/bioclients/pkg-r")
# or
install.packages("bioclients", repos = "https://samuelbharti.r-universe.dev")
```

## 2. One gene, one envelope

```{r}
res <- mygene_gene("TP53")
res$status
body_or_null(res)
```

It is a biohttp envelope. `status`, `error`, `detail`, `data`. Nothing new to
learn per service, which is the whole point of 29 of them sharing a
transport.

## 3. Nothing found is an answer

A symbol that resolves to nothing comes back `no_data`. The service was
reached and said so. That is different from the service being down, and the
status keeps the two apart.

```{r}
none <- mygene_gene("NOTAREALGENE123")
none$status
none$error
```

## 4. Batch: one row per input, in input order

The batch POST is the reason this client is worth installing. Three symbols
is one round trip, not three, and the round trip is where the time goes.

```{r}
batch <- mygene_genes(c("TP53", "NOTAREALGENE123", "BRAF"))
batch$status
batch$data[, c("symbol", "name", "entrez", "ensembl_gene", "uniprot")]
```

The middle row is `NA`, not missing. A table with two rows would silently
shift BRAF onto the second input, and every join downstream would be wrong.
Every batch function in the package keeps this contract.

## 5. The scoring trap

MyGene ranks hits by a score that can put an alias of a different gene above
the exact symbol. Query `TTN` with alias scopes on and TTR scores higher.
`mygene_pick_hit()` prefers an exact symbol match and only falls back to the
score when there is none.

```{r}
hits <- list(
  list(symbol = "TTR", entrezgene = "7276", `_score` = 19.07),
  list(symbol = "TTN", entrezgene = "7273", `_score` = 18.29)
)
mygene_pick_hit(hits, "TTN")$entrezgene
```

Traps like this are written down in the source with the date they were
verified, and pinned by a test. You inherit the fix instead of finding the
bug.

## 6. Client half, parser half

Every service module is two functions. The client builds the request, calls
biohttp and returns the envelope. The parser takes an already parsed body
and returns a tibble. The parser is pure, so it is tested offline against a
stored response body, and so can yours be.

```{r}
body <- jsonlite::fromJSON("fixtures/mygene_tp53.json", simplifyVector = FALSE)
mygene_parse_hits(body, "TP53")
```

`simplifyVector = FALSE` matters. The parsers expect the nested list shape
biohttp produces, not the data frames jsonlite guesses at.

```{r}
body <- jsonlite::fromJSON("fixtures/gnomad_constraint_braf.json", simplifyVector = FALSE)
gnomad_parse_constraint(body, "BRAF")
```

An empty body is `NULL`, which the client turns into `no_data`.

```{r}
reactome_parse_pathways(list())
```

The package ships 57 such fixtures under `tests/testthat/fixtures/`. The
whole suite runs without a network.

## 7. GraphQL, batched through aliases

gnomAD is GraphQL. Its limit is a query cost, not a request count, and the
verified ceiling is 25 aliases per query. The client chunks at 20 for
headroom and sends one request per chunk.

```{r}
transport_stats_reset()
cons <- gnomad_constraints(c("BRAF", "TP53", "EGFR"))
cons$data[, c("symbol", "pli", "loeuf", "mis_z")]
```

`transport_stats()` shows what the batched path sent. From the warmed cache,
nothing.

```{r}
transport_stats()
```

A GraphQL error arrives inside an HTTP 200, so a 2xx is not enough. Every
GraphQL client here runs `biohttp::graphql_error()` on the response before
parsing it.

## 8. Throttles are hard coded where they were verified

NCBI allows 3 requests per second, or 10 with an API key. VEP asks for 1 per
second. ClinGen was measured at 3. Each client carries its own default, so a
caller who has not read the service documentation still behaves.

```{r}
vep_default_throttle()
bioclients:::clinvar_throttle()
```

`clinvar_throttle()` is internal, which is the point: the ClinVar client
applies it by default and nobody has to remember the number.

The key itself is attached at dispatch through `secret_query`, which keeps
it out of the cache key and out of any printed request.

```r
clinvar_classification("NM_000546.6:c.524G>A")
# throttle = clinvar_throttle(), secret_query = list(api_key = NCBI_API_KEY)
```

None of those three is called in this document. They are slow by design,
and the point is made by the numbers.

## 9. Chaining sources

Open Targets takes an Ensembl id, not a symbol, because that is what its API
accepts. Resolve the symbol first. The client does not do that for you,
deliberately: a client that silently made a second call to a different
service would make a MyGene outage look like an Open Targets outage.

```{r}
acc <- res$data$uniprot
acc
dis <- uniprot_diseases(acc)
dis$status
dis$data[, c("name", "acronym", "mim", "causal")]
```

`causal` is read from the note text. "The disease is caused by variants
affecting the gene" is a Mendelian claim; "may be involved in the
pathogenesis" is not. Both arrive as disease comments, and collapsing them
would promote a hunch to a diagnosis.

Reactome takes the symbol directly and answers with a bare JSON array, which
a parser reaching for a `results` key would misread as "no pathways".

```{r}
paths <- reactome_pathways("NF1")
paths$status
head(paths$data[, c("pathway_id", "name", "in_disease")])
```

## 10. Testing a wrapper of your own

If you wrap a bioclients call in a function of your own, one mocked 503 is
enough to prove it passes the failure through instead of hiding it.

```{r}
httr2::with_mocked_responses(
  list(httr2::response(status_code = 503)),
  mygene_gene("BRCA1")$status
)
```

## What to take away

- 29 services, one envelope, one table shape per service, one row per input.
- Client half fetches, parser half is pure. Test the parser on a fixture.
- Batch functions never drop a row. A miss is an `NA` row in place.
- Verified limits and traps are in the source with dates. You inherit them.
- No Shiny dependency, on purpose. The app is the caller.

Docs: <https://www.samuelbharti.com/bioclients/>. Not on CRAN yet;
install from r-universe or with `pak::pak("samuelbharti/bioclients/pkg-r")`.
