---
title: "biohttp: a call returns a value"
subtitle: "One transport layer for R clients of biological web services"
---

```{r}
#| label: setup
#| message: false
source(here::here("R", "demo-env.R"))
library(biohttp)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(httr2_progress = FALSE)
packageVersion("biohttp")
```

biohttp knows how to make an HTTP call and report what happened. It does not
know what a gene is. Everything it does is in service of one rule:

**A call returns a value. It never raises.**

A DNS failure, a 503, a 200 carrying an HTML maintenance page, a host you
have already given up on: all four come back as a value you branch on. There
is no `tryCatch()` in this document, and there is none in the packages built
on top of it.

## 1. The soup this replaces

This is what a hand rolled client looks like in most Shiny apps. It is not a
strawman; it is a lightly edited copy of one I wrote.

```r
fetch_gene <- function(symbol) {
  url <- paste0("https://mygene.info/v3/query?q=", symbol)
  tryCatch(
    {
      body <- jsonlite::fromJSON(url)
      if (length(body$hits) == 0) return(NULL)
      body$hits[1, ]
    },
    error = function(e) NULL
  )
}
```

That `NULL` means four different things: the gene does not exist, the
network is down, MyGene returned a 503, or the body was not JSON. The caller
cannot tell them apart, so the UI says "not found" for all four. Worse, if
someone wraps this in a cache, the cache stores the `NULL` and keeps serving
"not found" after the outage ends.

## 2. The one rule

```{r}
res <- get_json(
  "https://mygene.info/v3",
  path = "gene/7157",
  query = list(fields = "symbol,name,taxid"),
  source = "MyGene"
)
res$ok
res$status
res$data$symbol
```

That call was served from the disk cache committed with this repo, which is
why it works with the network off. Section 9 shows the evidence.

## 3. What comes back

Every call returns the same fields, whatever happened.

```{r}
str(res[c("ok", "status", "http", "source", "error", "detail", "retry_after", "ts")])
```

| Field | What it holds |
|---|---|
| `ok` | `TRUE` only when `status` is `"ok"`. Derived, so the two cannot disagree. |
| `status` | One of seven levels. This is what you branch on. |
| `http` | The HTTP code, or `NA` when no response arrived. |
| `data` | The parsed body on success. |
| `source` | The label you passed, used in the message. |
| `error` | One sentence fit to show a user. |
| `detail` | The technical cause, for a log. Never shown to a user. |
| `retry_after` | Seconds the service asked you to wait, from a 429 or 503. |
| `ts` | When the envelope was built. |

`error` goes in the UI. `detail` goes in the log. A stack trace never reaches
a user, and the log never loses one.

## 4. The seven statuses, all of them, offline

```{r}
STATUS_LEVELS
```

`httr2::with_mocked_responses()` replaces the transport, so every HTTP
outcome can be produced on demand. `perform()` is the layer under
`get_json()`; it takes a request and a source label and returns an envelope.

```{r}
mock <- function(code) {
  httr2::with_mocked_responses(
    list(httr2::response(status_code = code)),
    perform(
      req_defaults(httr2::request("https://mock.test/x"), max_tries = 1),
      "Mock"
    )
  )
}
statuses <- rbind(
  data.frame(http = 200L, status = mock(200)$status),
  data.frame(http = 404L, status = mock(404)$status),
  data.frame(http = 429L, status = mock(429)$status),
  data.frame(http = 408L, status = mock(408)$status),
  data.frame(http = 503L, status = mock(503)$status)
)
statuses
```

Two more do not come from an HTTP code. `stale` is a value the client
decides on, for data it holds but no longer trusts:

```{r}
status_stale(data = list(n = 1), source = "Mock")$status
```

`skipped` means nothing was sent, because the host is paused. Section 8
produces one.

The three people miss:

- **`no_data`** is an answer, not a fault. The source was reached and has
  nothing for this query. A 404 from a lookup API means "no such record",
  which is information.
- **`skipped`** is not evidence about the query, only about the host.
- **`stale`** carries real data that is past its freshness window.

## 5. Branching

Handle the cases you can act on and let the rest fall through.

```{r}
render <- function(res) {
  switch(
    res$status,
    ok = paste("got", length(res$data), "fields"),
    no_data = "nothing found for that query",
    skipped = "source paused, try again shortly",
    rate_limited = paste("slow down, retry in", res$retry_after, "s"),
    res$error
  )
}
render(status_ok(data = list(a = 1, b = 2), source = "MyGene"))
render(status_no_data(source = "MyGene"))
render(status_error(source = "MyGene", http = 503L))
render(status_rate_limited(source = "MyGene", retry_after = 30))
```

If you genuinely do not care why a call failed, `body_or_null()` collapses
it. Reach for that when the failure is not actionable, not to avoid reading
the envelope.

```{r}
body_or_null(status_ok(data = list(n = 1)))
body_or_null(status_error(source = "MyGene"))
```

## 6. Retry needs a real server

Mocking replaces the transport below the retry loop, so a mocked 503 is
delivered once no matter what `max_tries` says. To show retry, run a real
local server. `webfakes` gives you one in a second process, so this is still
offline and still deterministic.

```{r}
#| label: fake-server
app <- webfakes::new_app()
app$locals$flaky <- 0L

# Fails twice, then succeeds.
app$get("/flaky", function(req, res) {
  req$app$locals$flaky <- req$app$locals$flaky + 1L
  n <- req$app$locals$flaky
  res$set_status(if (n < 3L) 503L else 200L)$send_json(
    list(attempts = n),
    auto_unbox = TRUE
  )
})
app$get("/count", function(req, res) {
  res$send_json(list(flaky = req$app$locals$flaky), auto_unbox = TRUE)
})
app$get("/slow", function(req, res) {
  Sys.sleep(3)
  res$send_json(list(ok = TRUE), auto_unbox = TRUE)
})
app$get("/limited", function(req, res) {
  res$set_status(429L)$set_header("Retry-After", "1")$send_json(
    list(error = "too many requests"),
    auto_unbox = TRUE
  )
})
app$get("/broken", function(req, res) {
  res$set_status(503L)$send_json(list(error = "down"), auto_unbox = TRUE)
})
fake <- webfakes::new_app_process(app)
fake$url()
```

```{r}
res <- perform(
  req_defaults(httr2::request(fake$url("/flaky")), max_tries = 4, timeout = 10),
  "Flaky"
)
res$status
res$data$attempts
```

Three attempts, one answer, no code at the call site.

## 7. Timeout and rate limit, live

`/slow` takes three seconds. With a one second budget, the envelope says
`timeout` and `http` is `NA`, because no response arrived.

```{r}
slow <- get_json(fake$url(), path = "slow", source = "Slow", timeout = 1, max_tries = 1)
slow$status
slow$http
slow$error
```

`/limited` answers 429 with a `Retry-After` header. The envelope reads it,
so the caller knows how long to wait instead of guessing.

```{r}
lim <- get_json(fake$url(), path = "limited", source = "Limited", max_tries = 1)
lim$status
lim$retry_after
```

In a batch, that header is honoured automatically: the retry pass waits the
time asked for before re-sending only the requests that were rate limited.
`retry_after` and the batched retry are biohttp 0.1.3; CRAN 0.1.2 accepts
`max_tries` on the batched functions and does nothing with it.

## 8. The circuit breaker

After three consecutive transport failures, a host is skipped for a cooldown
rather than waited on. Nothing listens on port 9, so each call fails at the
socket.

```{r}
dead <- "http://localhost:9"
for (i in 1:3) {
  r <- get_json(dead, path = "x", source = "Dead", timeout = 2, max_tries = 1)
  cat(i, r$status, "\n")
}
breaker_open("localhost")
```

The fourth call sends nothing. It comes back in microseconds with a status
that says so.

```{r}
r <- get_json(dead, path = "x", source = "Dead", timeout = 2, max_tries = 1)
r$status
r$error
```

The rule that matters, and the one a hand rolled breaker most often gets
wrong: **only a transport failure counts against a host**. A 503 proves the
host is reachable, so it must not open the breaker. A source returning error
pages is broken, but it is up, and taking it out of rotation would turn a
partial outage into a total one.

```{r}
b <- get_json(fake$url(), path = "broken", source = "Broken", max_tries = 1)
b$status
breaker_open("127.0.0.1")
```

The breaker heals after the cooldown on its own. `breaker_reset()` exists for
tests and for demos.

```{r}
breaker_reset()
breaker_open("localhost")
```

## 9. Caching, and what is never cached

Successes are cached. Failures never are. That second half is why the soup
in section 1 was dangerous: a cache that stores a failure poisons itself for
the life of the process.

The MyGene call in section 2 came from disk. Call it again: same data, and
no time spent waiting on a socket.

```{r}
again <- get_json(
  "https://mygene.info/v3",
  path = "gene/7157",
  query = list(fields = "symbol,name,taxid"),
  source = "MyGene"
)
identical(res$data, again$data)
system.time(get_json(
  "https://mygene.info/v3",
  path = "gene/7157",
  query = list(fields = "symbol,name,taxid"),
  source = "MyGene"
))[["elapsed"]]
```

A failure is fetched again every time. Count how often the mock is hit.

```{r}
hits <- 0L
always_503 <- function(req) {
  hits <<- hits + 1L
  httr2::response(status_code = 503)
}
httr2::with_mocked_responses(always_503, {
  first <- get_json("https://mock.test", path = "fail", source = "Mock", max_tries = 1)
  second <- get_json("https://mock.test", path = "fail", source = "Mock", max_tries = 1)
})
c(first$status, second$status)
hits
```

Two calls, two fetches. Had the 503 been cached, the second would have been
served without asking.

The cache is memory only by default. The disk tier is opt in, through
environment variables, because a library should not start writing to
somebody's disk because they installed it.

| Variable | Default | What it does |
|---|---|---|
| `BIOHTTP_CACHE_TTL` | 1800 | Lifetime of an entry, seconds |
| `BIOHTTP_CACHE_MAX_SIZE` | 256 MB | Memory tier ceiling, bytes |
| `BIOHTTP_CACHE_MAX_N` | unlimited | Memory tier ceiling, entries |
| `BIOHTTP_CACHE_DISK` | off | Set to `true` to add the disk tier |
| `BIOHTTP_CACHE_DIR` | `R_user_dir("biohttp", "cache")` | Disk tier location |
| `BIOHTTP_CACHE_DISK_TTL` | 7 days | Disk tier max age, seconds |
| `BIOHTTP_CACHE_SALT` | empty | Set per deployment |

This repo sets all of them in `R/demo-env.R`: the directory is inside the
project, both lifetimes are a year, and the salt is fixed. That is what makes
a committed cache replay. Per call, `ttl = 86400` keeps a bulk table for a
day without changing the default for everything else.

## 10. Many questions to one source

A gene list is many questions to one host. `get_json_many()` takes a list of
queries, serves what the cache holds, sends only the rest, and returns
envelopes in the order asked.

```{r}
transport_stats_reset()
many <- get_json_many(
  "https://mygene.info/v3",
  path = "query",
  queries = lapply(
    c("BRCA1", "TP53", "EGFR"),
    function(g) list(q = g, fields = "symbol,entrezgene", size = 1)
  ),
  source = "MyGene",
  throttle = list(capacity = 10, fill_time_s = 60)
)
vapply(many, function(r) r$status, character(1))
vapply(many, function(r) r$data$hits[[1]]$symbol, character(1))
```

`transport_stats()` reports what the batched path actually sent, per host.
Every one of those three was in the cache, so nothing went out.

```{r}
transport_stats()
```

Pass a `throttle`. httr2 will happily flood a host in parallel, and public
biological services run on a research budget. The mechanism is the package's;
the numbers are yours, because only you know what the service asks for.

## 11. Credentials

A token goes in a header, never in the URL. It is redacted from anything
printable and it stays out of access logs.

```{r}
req <- req_defaults(
  httr2::request("https://example.org/v1"),
  headers = list(Authorization = "Bearer a-real-token")
)
any(grepl("a-real-token", capture.output(print(req)), fixed = TRUE))
```

Some services insist on a key in the query string. `secret_query` attaches it
at dispatch only, so it is never in the cache key and never in a printed
request. bioclients uses it for `NCBI_API_KEY` on ClinVar.

```r
get_json(
  "https://eutils.ncbi.nlm.nih.gov/entrez/eutils",
  path = "esearch.fcgi",
  query = list(db = "clinvar", term = "BRCA1"),
  source = "ClinVar",
  secret_query = list(api_key = Sys.getenv("NCBI_API_KEY"))
)
```

## 12. Writing a client

A client is a thin function per endpoint: assemble, call, parse. There is no
error handling in it, because there is nothing to handle.

```{r}
mygene_query <- function(symbol, species = "human") {
  res <- get_json(
    "https://mygene.info/v3",
    path = "query",
    query = list(q = symbol, species = species),
    source = "MyGene"
  )
  if (!res$ok) {
    return(res)
  }
  status_ok(
    data = pluck_at(res$data, "hits", default = list()),
    source = "MyGene",
    http = res$http
  )
}
```

Two things to copy. Pass `source`, so the message says "MyGene is
temporarily unavailable" rather than "API is temporarily unavailable". And
return an envelope, not a bare value, so the caller can keep branching.

Test it offline. One mocked 503 is enough to prove the client passes a
failure through instead of hiding it.

```{r}
httr2::with_mocked_responses(
  list(httr2::response(status_code = 503)),
  mygene_query("BRCA1")$status
)
```

```{r}
#| include: false
fake$stop()
```

## What to take away

- One rule: a call returns a value. Branch on `status`, show `error`, log
  `detail`.
- Seven statuses. `no_data` and `skipped` are answers, not faults.
- Only a transport failure trips the breaker. A 500 proves the host is up.
- Successes are cached, failures never. With the disk tier and a fixed salt,
  a warmed cache replays a whole demo offline.
- Batch many questions to one host, pass a throttle, and read
  `transport_stats()` to see what actually went out.

Docs: <https://www.samuelbharti.com/biohttp/>. CRAN has 0.1.2; the batched
retry, `retry_after` and `transport_stats()` shown here are 0.1.3, from
`install.packages("biohttp", repos = "https://samuelbharti.r-universe.dev")`.
