---
title: "Validate, fetch, annotate"
subtitle: "Eleven gene symbols, one table, run twice, then with the plug pulled"
---

```{r}
#| label: setup
#| message: false
source(here::here("R", "demo-env.R"))
source(here::here("R", "demo-calls.R"))
library(biobouncer)
library(biohttp)
library(bioclients)
library(dplyr)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This is the whole pipeline the three packages exist for: a messy list of
gene symbols in, one annotated table out. It runs from the cache committed
with this repo, so it works with the network off, and the second half is
about proving that.

## 1. Time zero

Write down what you are running against before you run it.

```{r}
transport_stats_reset()
snap <- biobouncer_snapshots()
snap[snap$source == "hgnc", ]
c(biobouncer = as.character(packageVersion("biobouncer")),
  biohttp = as.character(packageVersion("biohttp")),
  bioclients = as.character(packageVersion("bioclients")))
```

## 2. The input

Eleven rows. Two retired symbols, a UniProt entry name pasted into a gene
column, one blank, one mixed case, and one that trips a scoring trap
downstream.

```{r}
genes <- read.csv("data/genes.csv", stringsAsFactors = FALSE, na.strings = "")
genes
```

## 3. Gate

`cache` mode checks every symbol against the dated HGNC snapshot. A retired
symbol gets its successor as a suggestion. `TP53_HUMAN` gets nothing, because
nothing in the snapshot is close enough to vouch for.

One caution before accepting suggestions blindly: in `cache` mode a near miss
such as `TP35` gets a fuzzy suggestion (`API5`), and a fuzzy match is a guess
for a human to review, not a repair. For this input every suggestion is a
retired-symbol successor, so the gate takes them.

```{r}
chk <- check_id(genes$symbol, "hgnc", how = "cache")
chk[, c("input", "valid", "normalized", "suggestion")]
genes$clean <- ifelse(chk$valid, chk$normalized, chk$suggestion)
genes$gate <- case_when(
  is.na(genes$symbol) ~ "missing",
  chk$valid ~ "valid",
  !is.na(chk$suggestion) ~ "repaired",
  TRUE ~ "rejected"
)
genes
```

Nine symbols go on. `TP53_HUMAN` and the blank stop here, before a single
request is spent on them.

```{r}
symbols <- genes$clean[!is.na(genes$clean)]
symbols
```

## 4. Resolve

One POST resolves all nine to Ensembl, Entrez and UniProt ids. `TTN` comes
back as TTN, not TTR, because the client prefers an exact symbol match over
MyGene's score.

```{r}
ids <- mygene_genes(symbols)
ids$status
ids$data[, c("symbol", "name", "entrez", "ensembl_gene", "uniprot")]
```

## 5. Annotate, batched

gnomAD constraint for all nine, through GraphQL aliases, in one request.

```{r}
cons <- gnomad_constraints(symbols)
cons$status
cons$data[, c("symbol", "pli", "loeuf", "mis_z")]
```

## 6. Annotate, one call per gene

Open Targets takes the Ensembl id. This is nine single calls, each its own
envelope, so a failure on one gene is one row of `NA` and a status you can
see, not an exception that stops the loop.

```{r}
ensembl <- ids$data$ensembl_gene
ot <- lapply(ensembl, function(id) opentargets_gene_diseases(id, size = 3))
top <- tibble(
  symbol = ids$data$symbol,
  ot_status = vapply(ot, function(r) r$status, character(1)),
  top_disease = vapply(ot, function(r) {
    if (isTRUE(r$ok)) r$data$disease[1] else NA_character_
  }, character(1)),
  ot_score = vapply(ot, function(r) {
    if (isTRUE(r$ok)) r$data$score[1] else NA_real_
  }, numeric(1))
)
top
```

## 7. One table

Join everything back onto the original eleven rows. The input order is kept
end to end, and the rows the gate rejected are still there, with their
reason, so the person who pasted `TP53_HUMAN` finds out.

```{r}
result <- genes |>
  left_join(
    ids$data |> select(clean = symbol, name, entrez, ensembl_gene),
    by = "clean"
  ) |>
  left_join(
    cons$data |> select(clean = symbol, pli, loeuf),
    by = "clean"
  ) |>
  left_join(top |> rename(clean = symbol), by = "clean")
result |> select(symbol, gate, clean, entrez, pli, loeuf, top_disease, ot_status)
```

## 8. What went over the wire

`transport_stats()` counts the batched path: MyGene and gnomAD. The nine
Open Targets calls went through the single call path, which the counter does
not see, so they are absent here rather than zero. From the warmed cache,
nothing was dispatched at all.

```{r}
transport_stats()
```

## 9. Run it again

Time the resolve and annotate steps on a second pass. Same tables, and the
whole thing takes a fraction of a second, because not one request leaves the
process.

```{r}
elapsed <- system.time({
  ids2 <- mygene_genes(symbols)
  cons2 <- gnomad_constraints(symbols)
  ot2 <- lapply(ensembl, function(id) opentargets_gene_diseases(id, size = 3))
})[["elapsed"]]
elapsed
identical(ids$data, ids2$data)
identical(cons$data, cons2$data)
transport_stats()
```

That is the reproducibility story in three lines. The cache stores only
successes, it is keyed on the exact request, and it lives in a directory you
can commit. Someone who clones this repo gets the same table without a
network.

## 10. Pull the plug

Now MyGene goes down. The breaker is tripped by hand here, three transport
failures recorded against the host, which is what three refused connections
would do on their own. The request has to be one the cache does not hold, so
one more gene is added to the list.

```{r}
for (i in 1:3) breaker_record("mygene.info", reachable = FALSE)
breaker_open("mygene.info")
down <- mygene_genes(c(symbols, "ABL1"))
down$status
down$error
```

Nothing was sent, nothing crashed, and the table still builds. The gene
columns from MyGene are `NA` for this run and the status column says why.

```{r}
resolved <- if (isTRUE(down$ok)) down$data else tibble(symbol = c(symbols, "ABL1"))
tibble(symbol = c(symbols, "ABL1")) |>
  left_join(resolved, by = "symbol") |>
  mutate(mygene_status = down$status) |>
  select(symbol, any_of(c("entrez", "ensembl_gene")), mygene_status)
```

The breaker heals after its cooldown. For the next section, reset it now.

```{r}
breaker_reset()
breaker_open("mygene.info")
```

## 11. Write it down

The methods paragraph writes itself from what the packages already know.

```{r}
cat(sprintf(
  "Symbols were validated against HGNC snapshot %s with biobouncer %s. Gene ids were resolved through MyGene.info and constraint scores fetched from gnomAD v4 through bioclients %s on biohttp %s. Responses were cached on %s.\n",
  snap$version[snap$source == "hgnc" & snap$version != "sample"][1],
  packageVersion("biobouncer"),
  packageVersion("bioclients"),
  packageVersion("biohttp"),
  format(file.mtime(list.files(cache_dir, full.names = TRUE))[1], "%Y-%m-%d")
))
```

```{r}
length(list.files(cache_dir, pattern = "\\.rds$"))
```

## What to take away

- Gate first. Two of eleven inputs never cost a request.
- Repair, do not drop. `MLL` and `CARS` became `KMT2A` and `CARS1` and went
  on to be annotated.
- One row per input, in input order, at every step, so the final join is
  trivial and honest.
- A second run is free and identical.
- An outage is a status in a column, not a crash.
